---
title: "Reproducible Research: Peer Assessment 1"
author: "Robert Deprizio"
date: "5/31/2018"
output: html_document
keep_md: true
---

## Loading and preprocessing the data

```{r preprocessing}
#setting my working directory
setwd("/Users/user/datasciencecoursera")

#Reading in data 
rawdata <- read.csv("data/activity.csv")
rawdata$date<-as.Date(rawdata$date)
```

## What is mean total number of steps taken per day?

```{r MeanSteps}
#Calculate the total steps taken per day 
total_steps<-aggregate(list(TotalSteps=rawdata$steps),by=list(Date=rawdata$date),sum,na.omit=TRUE)

#Calculate mean & median of total steps per day ignoring days with no data
total_steps$Mean<-tapply(rawdata$steps,rawdata$date,mean,na.omit=TRUE)
total_steps$Median<-tapply(rawdata$steps,rawdata$date,median,na.omit=TRUE)
total_steps<-na.omit(total_steps)
```

```{r plot1, echo=FALSE}
hist(total_steps$TotalSteps,main="Histogram of Total Steps Taken Each Day",
   xlab = "Number of Steps per Day",col="blue")
```
Reporting the mean and median for each day
```{r MeanSteps2}
print(total_steps)
```


## What is the average daily activity pattern?
```{r AvgDaily1}
#Calculate average steps taken for each interval 
#across all days excluding days without data
interval_means<-aggregate(list(Steps=rawdata$steps), list(Interval=rawdata$interval),  mean,na.rm = TRUE)
```
```{r plot2, echo=FALSE}
#Plot average steps taken on each interval across all days
plot(interval_means$Interval,interval_means$Steps,main="Average Daily Activity Pattern",xlab ="Time Interval",xaxt="n",xlim = c(0,2400), ylab="Average steps taken",type="l")
axis(side = 1, at=seq(0,2400,30), lwd.ticks=1)
```

Time interval with highest average and it's average
```{r AvgDaily2}
interval_means[interval_means$Steps==max(interval_means$Steps),]
```

## Imputing missing values
Calculating and reporting the total number of missing values in the dataset
```{r Imputting1}
#count number of missing records
sum(is.na(rawdata))
```

```{r Imputting2}
#Creating new data set that imputs the respective daily interval average, 
#(across all days) into intervals that had missing values 
rawdata2<-rawdata
rawdata2$steps<-ifelse(!is.na(rawdata2$steps), rawdata2$steps, interval_means$Steps)

#calculate the total steps taken per day for new data set
total_steps2<-aggregate(list(Steps=rawdata2$steps),by=list(Date=rawdata2$date),sum)
```
```{r plot3, echo=FALSE}
#Histogram of total steps taken each day for new data set
hist(total_steps2$Steps,main="Histogram of Total Steps Taken Each Day (New data set)",xlab = "Number of Steps per Day",col="red")
```



Reporting mean and median across all days for new data set
```{r Imputting3}
#Calculate mean & median of total steps per day for new data set
total_steps2$Mean<-tapply(rawdata2$steps,rawdata2$date,mean)
total_steps2$Median<-tapply(rawdata2$steps,rawdata2$date,median)
print(total_steps2)
```



## Are there differences in activity patterns between weekdays and weekends?

```{r weekdays1}
#creating 2 new variables that label 
#the day of the week and day type 
rawdata2$day_of_week<-weekdays(rawdata2$date)
rawdata2$day_type<-ifelse(rawdata2$day_of_week=="Saturday"|rawdata2$day_of_week=="Sunday","Weekend","Weekday")

#calculate the average steps taken per interval 
#across weekends and week days
avg_steps_days<-aggregate(steps ~ interval + day_type, data = rawdata2, mean)
```

```{r plot4, echo=FALSE}

#Plotting average steps per interval for weekdays and weekends 
plot(avg_steps_days[avg_steps_days$day_type=="Weekday",1],
     avg_steps_days[avg_steps_days$day_type=="Weekday",3],
     main="Average Daily Activity Pattern by interval (Weekdays)",
     xlab ="Time Interval",xaxt="n",xlim = c(0,2400), ylab="Average steps taken",type="l")
axis(side = 1, at=seq(0,2400,30), lwd.ticks=1)

plot(avg_steps_days[avg_steps_days$day_type=="Weekend",1],
     avg_steps_days[avg_steps_days$day_type=="Weekend",3],
     main="Average Daily Activity Pattern by interval (Weekends)",
     xlab ="Time Interval",xaxt="n",xlim = c(0,2400), ylab="Average steps taken",type="l")
axis(side = 1, at=seq(0,2400,30), lwd.ticks=1)
```
